---
title: "Exercise 3 - Ruhi Mahendra"
output:
  github_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(arrow)
```

## Load data

Load the following data: + applications from `app_data_sample.parquet` + edges from `edges_sample.csv`

```{r load-data}
# change to your own path!
data_path <- "~/Desktop/MMA/2023-ona-assignments/"
applications <- read_parquet(paste0(data_path,"app_data_sample.parquet"))
edges <- read_csv(paste0(data_path,"edges_sample.csv"))

applications
edges
```

## Get gender for examiners

We'll get gender based on the first name of the examiner, which is recorded in the field `examiner_name_first`. We'll use library `gender` for that, relying on a modified version of their own [example](https://cran.r-project.org/web/packages/gender/vignettes/predicting-gender.html).

Note that there are over 2 million records in the applications table -- that's because there are many records for each examiner, as many as the number of applications that examiner worked on during this time frame. Our first step therefore is to get all *unique* names in a separate list `examiner_names`. We will then guess gender for each one and will join this table back to the original dataset. So, let's get names without repetition:

```{r gender-1}
library(gender)
#install_genderdata_package() # only run this line the first time you use the package, to get data for it

# get a list of first names without repetitions
examiner_names <- applications %>% 
  distinct(examiner_name_first)

examiner_names
```

Now let's use function `gender()` as shown in the example for the package to attach a gender and probability to each name and put the results into the table `examiner_names_gender`

```{r gender-2}
# get a table of names and gender
examiner_names_gender <- examiner_names %>% 
  do(results = gender(.$examiner_name_first, method = "ssa")) %>% 
  unnest(cols = c(results), keep_empty = TRUE) %>% 
  select(
    examiner_name_first = name,
    gender,
    proportion_female
  )

examiner_names_gender
```

Finally, let's join that table back to our original applications data and discard the temporary tables we have just created to reduce clutter in our environment.

```{r gender-3}
# remove extra colums from the gender table
examiner_names_gender <- examiner_names_gender %>% 
  select(examiner_name_first, gender)

# joining gender back to the dataset
applications <- applications %>% 
  left_join(examiner_names_gender, by = "examiner_name_first")

# cleaning up
rm(examiner_names)
rm(examiner_names_gender)
gc()

```

## Guess the examiner's race

We'll now use package `wru` to estimate likely race of an examiner. Just like with gender, we'll get a list of unique names first, only now we are using surnames.

```{r race-1}
library(wru)

examiner_surnames <- applications %>% 
  select(surname = examiner_name_last) %>% 
  distinct()

examiner_surnames
```

We'll follow the instructions for the package outlined here <https://github.com/kosukeimai/wru>.

```{r race-2}
examiner_race <- predict_race(voter.file = examiner_surnames, surname.only = T) %>% 
  as_tibble()

examiner_race
```

As you can see, we get probabilities across five broad US Census categories: white, black, Hispanic, Asian and other. (Some of you may correctly point out that Hispanic is not a race category in the US Census, but these are the limitations of this package.)

Our final step here is to pick the race category that has the highest probability for each last name and then join the table back to the main applications table. See this example for comparing values across columns: <https://www.tidyverse.org/blog/2020/04/dplyr-1-0-0-rowwise/>. And this one for `case_when()` function: <https://dplyr.tidyverse.org/reference/case_when.html>.

```{r race-3}
examiner_race <- examiner_race %>% 
  mutate(max_race_p = pmax(pred.asi, pred.bla, pred.his, pred.oth, pred.whi)) %>% 
  mutate(race = case_when(
    max_race_p == pred.asi ~ "Asian",
    max_race_p == pred.bla ~ "black",
    max_race_p == pred.his ~ "Hispanic",
    max_race_p == pred.oth ~ "other",
    max_race_p == pred.whi ~ "white",
    TRUE ~ NA_character_
  ))

examiner_race
```

Let's join the data back to the applications table.

```{r race-4}
# removing extra columns
examiner_race <- examiner_race %>% 
  select(surname,race)

applications <- applications %>% 
  left_join(examiner_race, by = c("examiner_name_last" = "surname"))

rm(examiner_race)
rm(examiner_surnames)
gc()
```

## Examiner's tenure

To figure out the timespan for which we observe each examiner in the applications data, let's find the first and the last observed date for each examiner. We'll first get examiner IDs and application dates in a separate table, for ease of manipulation. We'll keep examiner ID (the field `examiner_id`), and earliest and latest dates for each application (`filing_date` and `appl_status_date` respectively). We'll use functions in package `lubridate` to work with date and time values.

```{r tenure-1}
library(lubridate) # to work with dates

examiner_dates <- applications %>% 
  select(examiner_id, filing_date, appl_status_date) 

examiner_dates
```

The dates look inconsistent in terms of formatting. Let's make them consistent. We'll create new variables `start_date` and `end_date`.

```{r tenure-2}
examiner_dates <- examiner_dates %>% 
  mutate(start_date = ymd(filing_date), end_date = as_date(dmy_hms(appl_status_date)))
```

Let's now identify the earliest and the latest date for each examiner and calculate the difference in days, which is their tenure in the organization.

```{r tenure-3}
examiner_dates <- examiner_dates %>% 
  group_by(examiner_id) %>% 
  summarise(
    earliest_date = min(start_date, na.rm = TRUE), 
    latest_date = max(end_date, na.rm = TRUE),
    tenure_days = interval(earliest_date, latest_date) %/% days(1)
    ) %>% 
  filter(year(latest_date)<2018)

examiner_dates
```

Joining back to the applications data.

```{r tenure-4}
applications <- applications %>% 
  left_join(examiner_dates, by = "examiner_id")

rm(examiner_dates)
gc()
```

Pick two workgroups you want to focus on (remember that a workgroup is represented by the first 3 digits of `examiner_art_unit` value) How do they compare on examiners' demographics? Show summary statistics and plots.

```{r demographics}
New  <- applications %>% select(examiner_art_unit, gender, race, tenure_days) 
New$examiner_art_unit = substr(New$examiner_art_unit, 1, 3) 

workgroup_171 <- New %>% filter(examiner_art_unit == 171)

workgroup_174 <- New %>% filter(examiner_art_unit == 174)
```

summary statistics workgroup_171

```{r 171}
library(vtable)
st(workgroup_171)
```


plot for workgroup_171

```{r plot 171}
ggplot(workgroup_171, aes(x = race, fill = gender)) +
  geom_bar(position = "dodge") +
  xlab("Race") +
  ylab("Count") +
  ggtitle("Distribution of Gender and Race in Workgroup 171")

```

summary statistics workgroup_174

```{r 174}
library(vtable)
st(workgroup_174)
```

plot for workgroup_174

```{r plot 174}
ggplot(workgroup_174, aes(x = race, fill = gender)) +
  geom_bar(position = "dodge") +
  xlab("Race") +
  ylab("Count") +
  ggtitle("Distribution of Gender and Race in Workgroup 174")

```

Create advice networks from `edges_sample` and calculate centrality scores for examiners in your selected workgroups

Create nodes

```{r node}
nodes <- applications %>%
  distinct(examiner_id) %>%
  select(examiner_id)
nodes
```

create edges

```{r edge}
applications$workgroups <- substr(applications$examiner_art_unit, 1, 3) 
examiner_workgroup <- subset(applications, select = c("examiner_id", "workgroups"))
edges$ego_examiner_workgroup <- examiner_workgroup$workgroups[match(edges$ego_examiner_id, examiner_workgroup$examiner_id)]
# Look up the workgroup corresponding to each alter_examiner_id in the examiner_workgroup dataframe
edges$alter_examiner_workgroup <- examiner_workgroup$workgroups[match(edges$alter_examiner_id, examiner_workgroup$examiner_id)]
# Filter out observations where either ego_examiner_workgroup or alter_examiner_workgroup is NA
edges_filtered <- edges[complete.cases(edges[, c("ego_examiner_id", "alter_examiner_id", "ego_examiner_workgroup", "alter_examiner_workgroup")]), ]
#Create the edges dataframe
edges_1 <- edges_filtered %>%
  distinct(ego_examiner_id, alter_examiner_id) %>%
  select(ego_examiner_id, alter_examiner_id)
edges_1
```

create network graph

```{r network}
library(igraph)
# Create network graph
g <- graph_from_data_frame(d = edges_1, directed = FALSE, vertices = nodes)

```

Get all centralities for network
```{r network centralities}
# Calculate the degree centrality of each node (examiner ID)
degree_centrality <- degree(g, mode = "all", normalized = FALSE)
betweeness_centrality <- betweenness(g)
closeness_centrality <- closeness(g, mode = "all", normalized = FALSE)
# Combine the centrality scores and node IDs into a table
centrality_table <- data.frame(examiner_id = V(g)$name, degree = as.vector(degree_centrality), betweeness = as.vector(betweeness_centrality),
closeness = as.vector(closeness_centrality))
                        
centrality_table$examiner_id <- as.numeric(centrality_table$examiner_id)
```

```{r degree_centrality 171}
New  <- applications %>% select(examiner_id,examiner_art_unit, gender, race, tenure_days, filing_date, patent_issue_date, abandon_date) 

all_centrality <- left_join(New, centrality_table, by = "examiner_id")

```


```{r app_proc_time}

all_centrality <- all_centrality %>% mutate(app_proc_time = 
  case_when(!is.na(patent_issue_date) ~ patent_issue_date - filing_date,
            !is.na(abandon_date) ~ abandon_date - filing_date
              )
)

```

``` {r model}
#degree
model_1 <- lm(app_proc_time ~ degree + race + gender + tenure_days, data = all_centrality)

#betweeness
model_2 <- lm(app_proc_time ~ betweeness + race + gender + tenure_days, data = all_centrality)

#closeness
model_3 <- lm(app_proc_time ~ closeness + race + gender + tenure_days, data = all_centrality)

model_1
model_2
model_3
```
Degree Centrality: Degree centrality increases processing time as the centrality is higher. Being black, Hispanic or white decreases the processing time. Being a male increases the processing time. 

Betweeness Centrality: Betweeness centrality increases processing time as the centrality is higher. Being black, Hispanic or white decreases the processing time. Being a male increases the processing time. 

Closeness Centrality: Closeness centrality decreases processing time as the centrality gets higher. Being black, Hispanic or white decreases the processing time. Being a male increases the processing time. 

``` {r model with gender}
#degree
model_4 <- lm(app_proc_time ~ degree*gender + race + tenure_days, data = all_centrality)

#betweeness
model_5 <- lm(app_proc_time ~ betweeness*gender + race + tenure_days, data = all_centrality)

#closeness
model_6 <- lm(app_proc_time ~ closeness*gender + race + tenure_days, data = all_centrality)

model_4
model_5
model_6

```
Degree Centrality: Degree centrality increases processing time as the centrality is higher. Being black, Hispanic or white decreases the processing time. Being a male increases the processing time. Having interactions between centrality and gender, lowers the processing time as centrality becomes higher and gender is male versus not having any interactions.

Betweeness Centrality: Betweeness centrality increases processing time as the centrality is higher. Being black, Hispanic or white decreases the processing time. Being a male increases the processing time. Having interactions between centrality and gender, increases the processing time as the centrality becomes higher versus not having any interactions.

Closeness Centrality: Closeness centrality decreases processing time as the centrality gets higher. Being black, Hispanic or white decreases the processing time. Being a male increases the processing time. Having interactions between centrality and gender increases the processing time as centrality becomes higher and gender is male versus not having any interactions.